---
title: "a short primer on JHPCE"
author: "Linda Gai"
date: "7/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using JHPCE

You will need to request a [JHPCE account](https://jhpce.jhu.edu/register/user/), attend a new user orientation with them, and take [HIPAA training](https://www.jhsph.edu/offices-and-services/institutional-review-board/training/index.html) before you can start using the cluster.

### Set up password-less authentication
Set this up [here](https://jhpce.jhu.edu/knowledge-base/authentication/ssh-key-setup/).

### Ensure you have X11 forwarding set up

X11 forwarding will allow you to send graphs to your laptop when you are using the cluster, so make sure you have it set up correctly. If you have a recent version of Mac OS X (newer than Mavericks), you may need to download XQuartz first.

If you get a warning when you sign into the cluster about X11 forwarding/xauth and/or still can't get graphs to show up while you're in the cluster, try troubleshooting using [this guide](https://stackoverflow.com/questions/27384725/ssh-x-warning-untrusted-x11-forwarding-setup-failed-xauth-key-data-not-gener/48543215#48543215).

### Accessing JHPCE from RStudio

To access the cluster from RStudio with passwordless authentication, send commands to the built-in Terminal on Mac OS X by holding option + command + Enter.

```{r,eval = FALSE}
ssh -X username@jhpce01.jhsph.edu
```

Make sure you don't get an error about X11 forwarding/xauth when you sign in (if you do, see the above section on X11 forwarding to fix it).

To use R on the cluster, access a computing node:
```{r,eval = FALSE}
#Allocate memory -- increase numbers if you run out of memory
qrsh -l mem_free=1G,h_vmem=1G,h_fsize=1G
```

Open R with the following commands:
```{r,eval = FALSE}
module load R
R
```

Now you can use R as normal. To check that X11 forwarding is working properly, run

```{r,eval = FALSE}
# Define the cars vector with 5 values
cars <- c(1, 3, 6, 4, 9)

# Graph the cars vector with all defaults
plot(cars)
```

A quartz window with a plot should appear on your laptop.

To get out of R, type

```{r,eval = FALSE}
q()
```
To sign out of the computing node, type
```{r,eval = FALSE}
logout
```
Make sure you sign out after you're done so other people can use the computing node!

To sign out of the login node, type
```{r,eval = FALSE}
logout
```

### More resources about using JHPCE

Jacob Fiksel wrote an excellent guide with many tips, tricks, and example shell scripts for using JHPCE efficiently on his Github [here](https://github.com/jfiksel/cluster-example). In the fall, he usually gives a presentation on using JHPCE as part of Computing Club. Try to go to this if you are able!

